``Interestingness'' in Mathematical Theory Exploration

### 2019-11-29 PhDThesis/bucketing.tex
Clustering text improvements

The results of this method are shown in Figure~\ref{fig:bucketing}, which plots
the difference between selecting clusters of definitions, compared to simply
choosing pseudorandomly (as a control). The results of clustering are
statistically indistinguishable from random.\footnote{Wilcoxon's signed-rank
  test finds no significant difference for most of the plotted points; whilst
  not a rigorous comparison (hence its ommission), this is nevertheless a clear
  indication of their similarity.} Despite this, two trends are apparent: a
slight increase in clustering's (relative) quality as the sample size increases,
and a decrease in variance as the number of subsets is increased. The variance
may be explained by the larger number of permutations available when using more
subsets, which decreases the chance of finding an outlier (either good \emph{or}
bad). The absolute increase in theorems found by clustering for large sizes, if
real, is dwarfed by the corresponding increase in the ground truth (the total
number of theorems available, without selecting subsets); the quality difference
between these selection methods as a \emph{proportion} of the ground truth
actually \emph{decreases} for larger sizes.

@@

We have demonstrated the effect of signature size (and hence the importance of
selection) on the time taken to explore, in particular noting the ``heavy tail''
of explorations timing out, which increases linearly with signature size. We
have attributed this to ``toxic'' definitions, whose presence in a signature
will exhaust the allocated time and/or memory. To mitigate this problem, and
hence reduce the severity of signature selection, we recommend a more
fine-grained approach to resource management and/or scheduling: by limiting the
resources of individual tests, evaluation of toxic definitions can be terminated
without having to abort the whole exploration run.

We have shown the impact of signature selection on the quality of discoverable
properties, by sampling sets of definitions from the Theory Exploration
Benchmark and comparing how many of their associated theorems become impossible
to discover (due to their definitions being separated) as these sets are
subdivided by different selection methods. Optimal selection (calculated with
knowledge of the benchmark theorems) retains a large proportion of the desired
output, showing the potential to maintain high quality. Pseudorandom selection
is found to be closer to pessimal, and selecting based on similarity (determined
via an unsupervised machine learning algorithm) performs no different to random.

Automated signature selection is hence an important problem, requiring further
research to establish if, and how, the potential benefits can be reaped.

@@

%\section{Experiments}

% TODO: This was taken from the Background section, so probably doesn't fit here
% without some fiddling.
% TODO: Describe all of our experimental setups:
%
%  - The use of the TEBenchmark methodology for running QSpec, including the
%    sampling process
%
%  - Running our recurrent signature selection algorithm on samples from TEBenchmark,
%    including how we picked the sizes
%
%  - The definition of HashSpec and its use as control; how we ran all of the
%    same samples through it
%
%  - The use of optimal and pessimal oracles, defined via constraint
%    satisfaction. How we it was only feasible to run these up to size 10 (11?)

%We follow the tradition of prior practitioners and use an existing corpus of
%theorems as a \emph{ground truth} against which to compare results. To analyse
%behaviour on large inputs (i.e. those for which exploration is currently
%infeasible), as required to study the effects of signature selection, we have
%used the Theory Exploration Benchmark of section~\ref{sec:tebenchmark}, which is
%based on the Tons of Inductive Problems (TIP) problem set for theorem
%provers~\cite{claessen2015tip}.

>>>>

The results of this method are shown in Figure~\ref{fig:bucketing}, which plots
the difference between selecting clusters of definitions, compared to simply
choosing pseudorandomly (as a control). The results of clustering are
statistically indistinguishable from random.\footnote{Wilcoxon's signed-rank
  test finds no significant difference for most of the plotted points; whilst
  not a rigorous overall comparison (hence its ommission), this is nevertheless
  a clear indication of their similarity.} Despite this, two trends are
apparent: a slight increase in clustering's (relative) quality as the sample
size increases, and a decrease in variance as the number of subsets is
increased. The variance may be explained by the larger number of permutations
available when using more subsets, which decreases the chance of finding an
outlier (either good \emph{or} bad). The absolute increase in theorems found by
clustering for large sizes, if real, is dwarfed by the corresponding increase in
the ground truth (the total number of theorems available, without selecting
subsets); the quality difference between these selection methods as a
\emph{proportion} of the ground truth actually \emph{decreases} for larger
sizes.

@@

We have demonstrated the effect of signature size (and hence the importance of
selection) on the time taken to explore, in particular noting the ``heavy tail''
of explorations timing out, which increases linearly with signature size. We
have attributed this to ``toxic'' definitions, whose presence in a signature
will exhaust the allocated time and/or memory. To mitigate this problem, and
hence the severity of signature selection's pruning, we recommend a more
fine-grained approach to resource management and/or scheduling: by limiting the
resources of individual tests, evaluation of toxic definitions can be terminated
without having to abort the whole exploration run.

We have shown the impact of signature selection on the quality of theory
exploration output, by sampling sets of definitions from the Theory Exploration
Benchmark and comparing how many of their associated theorems become impossible
to discover (due to their definitions being separated) as these sets are
subdivided by different selection methods. Optimal selection (calculated with
knowledge of the benchmark theorems) retains a large proportion of the desired
output, showing the potential to maintain high quality. Pseudorandom selection
is found to be closer to pessimal, and selecting based on similarity (determined
via an unsupervised machine learning algorithm) performs no different than
random.

Automated signature selection is hence an important unsolved problem, requiring
further research to establish if, and how, the potential benefits shown by our
oracle-assisted algorithms can be reaped.

@@

### 2019-11-29 PhDThesis/clustering.tex
Clustering text improvements

\iffalse TODO: Alison: Each chapter should have a description at the start (see
bucketing ``in this chapter'') \fi

In this chapter we consider how the contents of a given signature can be
analysed, such that subsequent exploration can be performed more efficiently. We
frame this analysis as a task of \emph{unsupervised machine learning}, and
describe how well-known \emph{clustering algorithms} can utilise the syntax of
definitions to determine their similarity. This requires a detailed description
of Haskell definition syntax, in particular of the ``Core'' intermediate
language of the GHC compiler. We identify the difficulty posed by references,
and how these can be made less opaque through the use of \emph{recurrent
  clustering}.

@@

More speculative directions include the use of \emph{learned} representations,
rather than our hand-crafted features \cite{bengio2013representation}. This
would provide an interesting comparison, as well as being more robust in the
face of language evolution. Another intriguing possibility is to extend the
recurrent nature of our algorithm to make use of the discovered properties
during clustering; for example, by treating the discovered equations as rewrite
rules to reduce the ASTs prior to feature extraction.
\fi
W

>>>>

In this chapter we consider how the contents of a given signature can be
analysed syntactically, to aid theory exploration systems (either by improving
efficiency, or finding more interesting properties). Inspired by the approaches
of \aclml{} and \mlforpg{}, we frame this analysis as a task of
\emph{unsupervised machine learning}, and describe how well-known
\emph{clustering algorithms} can utilise the syntax of definitions to determine
their similarity. This requires a detailed description of Haskell syntax, in
particular we use the ``Core'' intermediate language of the GHC compiler. We
identify the difficulty posed by references, and how these can be made less
opaque through the use of \emph{recurrent clustering}.

Our main contribution is a recurrent clustering method to determine the
syntactic similarity of Haskell definitions, which is transitive across
references (e.g. the similarity between definitions \hs{A} and \hs{B}, which
reference definitions \hs{C} and \hs{D} respectively, is correlated to the
similarity between \hs{C} and \hs{D}). We also provide an iterative
algorithm, which is more efficient than a straightforward (na\"ive)
implementation of the defined recurrence relation. We have made a prototype of
our system available, including a plugin for the \textsc{GHC} compiler to
enable end-to-end automation, taking a standard Haskell package and producing
clusters of definitions; these may subsequently be explored, e.g. via
\quickspec{}, which we investigate further in \S~\ref{sec:signatureselection}.

Without an independent ``gold standard'' to compare our results against, we
instead perform a qualitative analysis; using (translations of) examples from
\aclml{} and real Haskell packages.

@@

More speculative directions include the use of \emph{learned} representations,
rather than our hand-crafted features \cite{bengio2013representation}. This
would provide an interesting comparison, as well as being more robust in the
face of language evolution. Another intriguing possibility is to extend the
recurrent nature of our algorithm to make use of the discovered properties
during clustering; for example, by treating the discovered equations as rewrite
rules to reduce the ASTs prior to feature extraction.
\fi

% FIXME
